### 2016-07-04 case-studies.tex
Case studies changed significantly.

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a Postgres database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rc}
isolation of Postgres may lead to anomalous executions as confirmed by
our experiments; with 16 (resp.  32) concurrent writers performing
single increments, we observed a total of 7 (17) violations of
monotonicity. We subsequently instantiated our reasoning framework for
an {\sc sc} store with Postgres isolation levels (which are stronger
than ANSI SQL's and MySQL's), and spent 5 man hours determining
appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
formally reason about scenarios described in this section can be found
in the supplementary.} In particular, we were able to formally prove
that to preserve the invariant, writers need to execute at
\iso{Repeatable Read} isolation, while readers can continue at {\sc
rc}. Repeating the experiments with this configuration led to no
violations, as expected. While executing all transactions at {\sc ser}
also prevents violations, it led to an additional 47\% increase in
latency on an average (16 concurrent writers; 5 rounds). We repeated
the exercise for MySQL, and found that its {\sc rr} is not sufficient
to enforce monotonicity; {\sc ser} is needed.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
re-implemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The trade-off is that the view count is now no
longer accurate, although it grows monotonically.}.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behaviour is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, chosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. We then instantiated our framework for an
{\sc sc} store with various ANSI SQL isolation levels for user
deletions and timeline reads, and obtained a proof that {\sc rc} is
sufficient for deletions, whereas reads require {\sc rr}. The proof is
immediately applicable for MySQL, which implements ANSI SQL isolation
levels. Since Postgres's isolation levels are stronger than the
corresponding ANSI levels, \rulelabel{RG-Conseq} rule
(Fig.~\ref{fig:rg-rules}) lets us deduce that the same choice of
isolation levels also work on Postgres. Indeed, experiments confirmed
this judgment.

>>>>

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rr}
isolation of MySQL may lead to anomalous executions as confirmed by
our experiments; with 32 concurrent writers performing single
increments, and one reader issuing a constant stream of reads, we
observed a maximum of 12 violations of monotonicity in 10 rounds of
operation. We subsequently instantiated our reasoning framework for an
{\sc sc} store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
formally reason about scenarios described in this section can be found
in the supplementary.} In particular, we were able to formally prove
that to preserve the invariant, writers need to execute at
\iso{Serializable} isolation, while readers can execute at \iso{Read
Uncommitted}. Repeating the experiments with this configuration led to
no violations, as expected. While executing all transactions at {\sc
ser} also prevents violations, it led to an additional 113\% increase
in the latency of writes on an average (32 concurrent writers; 5
rounds). We repeated the experiments with the same configuration on
Postgres, and found no violations. However, Postgres's isolation
levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
ways. We therefore formalized Postgres's guarantees and repeated
relevant parts of the proof to discover that \iso{Repeatable Read} is
in fact sufficient for writers - an observation we confirmed through
experiments. As usual, using {\sc ser} for all transactions also
suffices, but on Postgres it led to an additional 47\% increase in
write latency.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
re-implemented the counter as a op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident as we
considered an instantiation our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
In other words, if the reads chose the largest valued write effect,
instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}.
We implemented this version of the counter as an op-based RDT, which
uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
latency compared to the version that uses serializable increments.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behavior is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, chosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. To determine appropriate isolation levels,
we then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expectected), while yeilding a 38\% better overall latency than {\sc
ser} configuration (96 concurrent clients; 5 rounds).  Since
Postgres's isolation levels are stronger than the corresponding ANSI
levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
deduce that the same choice of isolation levels also work on Postgres.
Indeed, experiments confirm this judgment.

### 2016-07-04 opsem.tex
Case studies changed significantly.

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

>>>>

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behavior is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behavior as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behavior of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

### 2016-07-04 related-work.tex
Case studies changed significantly.

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specified several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
to specify isolation levels with atomic visibility, but their
specifications are also for complete traces; Moreover, their
formulation implicitly assumes homogeneity, i.e., they require all
transactions to execute using either {\sc si} or {\sc ser}, which is
often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and different
forms of weak consistency, in contrast to this work, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

>>>>

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specified several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for
reasoning about program behavior. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
to specify isolation levels with atomic visibility, but their
specifications are also for complete traces; Moreover, their
formulation implicitly assumes homogeneity, i.e., they require all
transactions to execute using either {\sc si} or {\sc ser}, which is
often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and different
forms of weak consistency, in contrast to this work, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

### 2016-07-04 store-consistency.tex
Case studies changed significantly.

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicking the
behaviour of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

@@

%% \begin{remark}
%% Note that the purpose of maximum visibility principle is to
%% rationalize the observed behaviour of databases in practice. In
%% particular, it is not intended to be a guiding principle to engineer
%% data stores.
%% \end{remark}

>>>>

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicking the
behavior of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

@@

%% \begin{remark}
%% Note that the purpose of maximum visibility principle is to
%% rationalize the observed behavior of databases in practice. In
%% particular, it is not intended to be a guiding principle to engineer
%% data stores.
%% \end{remark}
